---
title: "Final Project Dashboard"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: scroll
---

```{r global, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(sf)
library(tigris)
library(leaflet)
library(censusapi)
library(mapview)
library(readxl)
library(RColorBrewer)
library(plotly)

Sys.setenv(CENSUS_KEY="67343486046933b2d74f13923ac153228348f817")

# temp <- tempfile()
# download.file("https://oehha.ca.gov/media/downloads/calenviroscreen/document/calenviroscreen40resultsdatadictionaryf2021.zip",destfile = temp)
# 
# ces4 <- read_excel("calenviroscreen40resultsdatadictionary_F_2021.xlsx", sheet = "CES4.0FINAL_results")
ces4 <- readRDS("ces4.rds")

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

cv_county_names <-
  c(
    "Butte", "Colusa", "Glenn", "Fresno", "Kern", "Kings", "Madera", "Merced", "Placer", "San Joaquin", "Sacramento", "Shasta", "Solano", "Stanislaus", "Sutter", "Tehama", "Tulare", "Yolo", "Yuba"
  )

ca_counties <- counties("CA", cb = T, progress_bar = F)

ca_tracts <- tracts("CA", cb = T, progress_bar = F)

census_race_categories <- 
  c(
    "White Alone",
    "Black or African American",
    "American Indian and Alaska Native Alone",
    "Asian Alone",
    "Native Hawaiian and Other Pacific Islander Alone",
    "Some Other Race Alone",
    "Two or More Races"
  )

bay_race_tracts <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "tract:*",
      regionin = "state:06+county:001,013,041,055,075,081,085,095,097",
      vars = paste0("B19001",LETTERS[x],"_001E")
    ) %>%
      mutate(
        tract = paste0(state, county, tract) %>% as.numeric(),
        race = census_race_categories[x]
      ) %>% 
      select(
        tract,
        race,
        estimate = paste0("B19001",LETTERS[x],"_001E")
      )
  })

cv_race_tracts <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "tract:*",
      regionin = "state:06+county:007,011,021,019,029,031,039,047,061,067,077,089,095,099,101,103,107,113,115",
      vars = paste0("B19001",LETTERS[x],"_001E")
    ) %>%
      mutate(
        tract = paste0(state, county, tract) %>% as.numeric(),
        race = census_race_categories[x]
      ) %>% 
      select(
        tract,
        race,
        estimate = paste0("B19001",LETTERS[x],"_001E")
      )
  })

ca_gama <- read.table("gama_usgs_statewide.txt", sep = "\t", header = T)
```

Inputs {.sidebar}
-------------------------------------

```{r}
selectInput(
  inputId = "region", 
  label = "CES4 Region",
  choices = c("Bay Area", "Central Valley"), 
  selected = "Bay Area"
)
```

```{r}
selectInput(
  inputId = "time", 
  label = "GAMA Well Data Time Frame:",
  # choices = c("2005-2010", "2010-2015", "2015-Present", "2018-Present", "2019-Present"),
  choices = c("2015-2017", "2017-2019", "2019-Present"),
  selected = "2019-Present"
)
```

```{r}
selectInput(
  inputId = "chemical", 
  label = "GAMA Groundwater Contaminant:",
  choices = c("Ammonia", "Benzene", "Cadmium", "Chromium", "Lead", "Naphthalene", "Nitrate", "Nitrite", "Orthophosphate", "Toluene", "Uranium", "Xylenes"), 
  selected = "Nitrate"
)
```

Column
-------------------------------------

### Introduction

Groundwater is an essential resource that provides water for a plethora of uses in urban systems, and one of these uses is for drinking. There are many issues associated with groundwater that are pertinent to the Bay Area, including groundwater depletion and contamination. There are many common groundwater contaminants, such as nitrate, lead, and volatile organic compounds, each of which can have severe detrimental health effects if found in drinking water. The issue of groundwater is a growing one in many regions of California so this project aims to explore the extent of this issue in the Bay Area by comparing groundwater contamination levels to other regions in the state. In many regions in San Francisco and the rest of the Bay Area, toxic waste, from industrial and agricultural processes, has been reported to have come into contact with groundwater reservoirs, which may lead to noticeable increases in contaminant concentrations in the future. The severity of groundwater-related issues is enhanced by factors such as flooding, storms, and in the case of coastal cities like San Francisco, sea level rise. Moreover, each of these factors will continue to grow in frequency and magnitude as climate change progresses, thus making the issue of understanding groundwater contamination that much more urgent.

CalEnviroScreen 4.0 report's aggregated indicators are useful tools for looking at a summarized overview of groundwater quality trends in California. Assessing the overall health of a groundwater source involves numerous factors primarily involving agricultural, industrial and commercial use of potentially hazardous materials. As a simplified assessment of the relative conditions for the groundwater sources in California, the CalEnviroScreen 4.0 (CES4) created the "Groundwater Threats" indicator as a measurement of "The nature and the magnitude of the threat and burden posed by sites maintained in GeoTracker vary significantly by site type (e.g., leaking underground storage tank or cleanup site) and status (e.g., Completed Case Closed or Active Cleanup)." Specifically, the indicator combines the site type and status data with the overall proximity to populated regions. Since the CES4 reports data at the Census Tract level, the Groundwater Threat score reported is the sum of the weighted scores for all the sites within the boundaries of the census tract.

### CalEnviroScreen 4.0 Groundwater Threat Map

```{r}
leafletOutput("map")
```

```{r, context = "server"}
observeEvent({
  input$region
  }, {

  county_names <- switch(
    input$region,
    "Bay Area" = bay_county_names,
    "Central Valley" = cv_county_names
  )

  ces4_gwthreats <-
    ces4 %>%
    filter(`California County` %in% county_names) %>%
    select(`Census Tract`, `Groundwater Threats`) %>%
    left_join(
      ca_tracts %>%
        transmute(GEOID = as.numeric(GEOID)),
      by = c("Census Tract" = "GEOID")
    ) %>%
    st_as_sf()

  gwthreats_pal <- colorNumeric(
    palette = "YlOrRd",
    domain = ces4_gwthreats$`Groundwater Threats`
  )

  output$map <- renderLeaflet({
    leaflet() %>%
      addProviderTiles(providers$CartoDB.Positron) %>%
      addPolygons(
        data = ces4_gwthreats,
        fillColor = ~gwthreats_pal(`Groundwater Threats`),
        color = "white",
        weight = 0.5,
        fillOpacity = 0.5,
        label = ~paste0("Census Tract: ", `Census Tract`, ", Groundwater Threat Score: ",`Groundwater Threats`)
      ) %>%
      addLegend(
        data = ces4_gwthreats,
        pal = gwthreats_pal,
        values = ~`Groundwater Threats`,
        title = paste0("Groundwater Threat Score in the ", input$region)
      )
  })

})
```

```{r}
# Commands used to calculate statistics
# quantile(ces4_bay_gwthreats$`Groundwater Threats`, probs = seq(.1, .9, by = .1))
# quantile(ces4_cv_gwthreats$`Groundwater Threats`, probs = seq(.1, .9, by = .1))
# sorted_bay_gwthreats <- sort(ces4_bay_gwthreats$`Groundwater Threats`, decreasing = TRUE)
# sorted_cv_gwthreats <- sort(ces4_cv_gwthreats$`Groundwater Threats`, decreasing = TRUE)
```

### Map and Histogram Comments

This map shows the Groundwater Threat score calculated for each census tract in the counties of the selected region. It is immediately evident that the highest concentration of high groundwater threat score tracts are located in the regions directly adjacent to the San Francisco Bay. For the Bay Area, the median score was 14.00 and the 90th percentile score was 66.25 for the 1581 tracts. As a result, it is quite alarming to see that the highest scores were 533, 478, and 378, all three of which geographically clustered in the area near the San Francisco-Oakland Bay Bridge. The maximum score belonged to Census Tract 4287 in Alameda, which has a population of 4,472 and is 27% White, 17% Black, 33% Asian, and 16% Hispanic (it should be noted that race statistics have a margin of error of at least 10 percent of the total value). The second highest score belonged to Census Tract 9806 in San Fransisco's Hunters Point neighborhood, which has a population of 690 and is 14% White, 34% Black, 25% Asian, and 21% Hispanic.

In comparison with the Bay Area, the Central Valley has an overall similarly shaped distribution of Groundwater Threat scores (which can be viewed in the histogram below). However, the quantile scores in the Central Valley were lower, with the median score being 5.75 and the 90th percentile score was 44.30 for the 1454 tracts. The maximum score of 513.75 belonged to Census Tract 4287 in Sacramento, which had overall low CES4 scores in other categories (less risk) while being in the 100th percentile for Groundwater Threat.

### CalEnviroScreen 4.0 Groundwater Threat Histogram

```{r}
plotlyOutput("histo")
```

```{r, context = "server"}
observeEvent({
  input$region
  }, {
    
  county_names <- switch(
    input$region,
    "Bay Area" = bay_county_names,
    "Central Valley" = cv_county_names
  )
    
  ces4_gwthreats <-
    ces4 %>%
    filter(`California County` %in% county_names) %>%
    select(`Census Tract`, `Groundwater Threats`) %>%
    left_join(
      ca_tracts %>%
        transmute(GEOID = as.numeric(GEOID)),
      by = c("Census Tract" = "GEOID")
    ) %>%
    st_as_sf()
    
  output$histo <- renderPlotly({
  gwthreats_histo <-
    plot_ly(
      x = ces4_gwthreats$`Groundwater Threats`,
      type = "histogram",
      marker = list(color = "#e9c46a")
    ) %>% 
    layout(
      xaxis = list(
        title = "Groundwater Threat Score",
        fixedrange = T
      ),
      yaxis = list(
        title = "Frequency (Census Tracts)",
        fixedrange = T
      ),
      title = list(
        text = paste0("Groundwater Threat Score Frequency in the ", input$region)
      )
    ) %>% 
    config(displayModeBar = F)
  })
})
```

### CalEnviroScreen 4.0 Groundwater Stacked Equity Chart

```{r}
plotlyOutput("equityStacked")
```

```{r, context = "server"}
observeEvent({
  input$region
  }, {

  race_tracts <-
    switch(
      input$region,
      "Bay Area" = bay_race_tracts,
      "Central Valley" = cv_race_tracts
    )

  county_names <- switch(
    input$region,
    "Bay Area" = bay_county_names,
    "Central Valley" = cv_county_names
  )

  ces4_gwthreats <-
    ces4 %>%
    filter(`California County` %in% county_names) %>%
    select(`Census Tract`, `Groundwater Threats`) %>%
    left_join(
      ca_tracts %>%
        transmute(GEOID = as.numeric(GEOID)),
      by = c("Census Tract" = "GEOID")
    ) %>%
    st_as_sf()
  
  gwthreats_race <-
    race_tracts %>% 
    left_join(
      ces4_gwthreats %>% 
        st_drop_geometry(),
      by = c("tract" = "Census Tract")
    ) %>% 
    mutate(
      gwthreats_tier =
        case_when(
          `Groundwater Threats` <= 1 ~ "0-1",
          `Groundwater Threats` <= 2 ~ "1-2",
          `Groundwater Threats` <= 5 ~ "2-5",
          `Groundwater Threats` <= 20 ~ "5-20",
          `Groundwater Threats` <= 50 ~ "20-50",
          `Groundwater Threats` <= 100 ~ "50-100",
          TRUE ~ "100-550"
        ) 
    ) %>% 
    group_by(race, gwthreats_tier) %>% 
    summarize(estimate = sum(estimate, na.rm = T))

  output$equityStacked <- renderPlotly({
    gwthreats_race_stacked <-
      gwthreats_race %>%
      ggplot() +
      geom_bar(
        aes(
          x = gwthreats_tier %>% factor(levels = rev(c("0-1","1-2","2-5","5-20","20-50","50-100","100-550"))),
          y = estimate,
          fill = race %>% factor(levels = rev(census_race_categories))
        ),
        stat = "identity",
        position = "stack"
      ) +
      scale_fill_manual(values = rev(c("#f94144", "#f3722c", "#f9c74f", "#43aa8b", "#33a1fd", "#344CB7", "#7267CB"))) +
      labs(
        x = "Groundwater Threat Score",
        y = "Number of Households",
        title = paste0(input$region, " Groundwater Threat Score by Race"),
        subtitle = "Sum of weighted scores for sites within each census tract,\naggregated by Census tract",
        fill = "Race of householder"
      ) +
      coord_flip() +
      theme(
        legend.position = "bottom",
        legend.direction = "vertical"
      ) +
      guides(
        fill = guide_legend(
          reverse = T
        )
      )
  })
})
```

### CalEnviroScreen 4.0 Groundwater Filled Equity Chart

```{r}
plotlyOutput("equityFilled")
```

```{r, context = "server"}
observeEvent({
  input$region
  }, {

  race_tracts <-
    switch(
      input$region,
      "Bay Area" = bay_race_tracts,
      "Central Valley" = cv_race_tracts
    )

  county_names <- switch(
    input$region,
    "Bay Area" = bay_county_names,
    "Central Valley" = cv_county_names
  )

  ces4_gwthreats <-
    ces4 %>%
    filter(`California County` %in% county_names) %>%
    select(`Census Tract`, `Groundwater Threats`) %>%
    left_join(
      ca_tracts %>%
        transmute(GEOID = as.numeric(GEOID)),
      by = c("Census Tract" = "GEOID")
    ) %>%
    st_as_sf()
  
  gwthreats_race <-
    race_tracts %>% 
    left_join(
      ces4_gwthreats %>% 
        st_drop_geometry(),
      by = c("tract" = "Census Tract")
    ) %>% 
    mutate(
      gwthreats_tier =
        case_when(
          `Groundwater Threats` <= 1 ~ "0-1",
          `Groundwater Threats` <= 2 ~ "1-2",
          `Groundwater Threats` <= 5 ~ "2-5",
          `Groundwater Threats` <= 20 ~ "5-20",
          `Groundwater Threats` <= 50 ~ "20-50",
          `Groundwater Threats` <= 100 ~ "50-100",
          TRUE ~ "100-550"
        ) 
    ) %>% 
    group_by(race, gwthreats_tier) %>% 
    summarize(estimate = sum(estimate, na.rm = T))

  output$equityFilled <- renderPlotly({
      gwthreats_race_fill <-
        gwthreats_race %>% 
        group_by(gwthreats_tier, race) %>% 
        summarize(estimate = sum(estimate)) %>% 
        rbind(
          gwthreats_race %>% 
            group_by(race) %>% 
            summarize(estimate = sum(estimate)) %>% 
            mutate(gwthreats_tier = "Total")
        ) %>% 
        ggplot() +
        geom_bar(
          aes(
            x = gwthreats_tier %>% factor(levels = rev(c("Total","0-1","1-2","2-5","5-20","20-50","50-100","100-550"))),
            y = estimate,
            fill = race %>% factor(levels = rev(census_race_categories))
          ),
          stat = "identity",
          position = "fill"
        ) +
        scale_fill_manual(values = rev(c("#f94144", "#f3722c", "#f9c74f", "#43aa8b", "#33a1fd", "#344CB7", "#7267CB"))) +
        labs(
          x = "Groundwater Threat Score",
          y = "Proportion of Households",
          title = paste0(input$region, " Groundwater Threat Score by Race"),
          subtitle = "Sum of weighted scores for sites within each census tract,\naggregated by Census tract",
          fill = "Race of householder"
        ) +
        coord_flip() +
        theme(
          legend.position = "bottom",
          legend.direction = "vertical"
        ) +
        guides(
          fill = guide_legend(
            reverse = T
        )
      )
  })
})
```

```{r include=FALSE}
# percent of population white
# ((sum(bay_race_total$estimate[7])/sum(bay_race_total$estimate))*100) %>% round()
# # percent asian
# ((sum(bay_race_total$estimate[2])/sum(bay_race_total$estimate))*100) %>% round()
# # percent black
# ((sum(bay_race_total$estimate[3])/sum(bay_race_total$estimate))*100) %>% round()
# # percent white in 0-1 and 1-2 tiers
# ((bay_gwthreats_race %>% 
#   filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[1:2]) %>% 
#   filter(race == "White Alone") %>% 
#   pull(estimate) %>% 
#   sum()) /
#   (bay_gwthreats_race %>% 
#     filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[1:2]) %>% 
#     pull(estimate) %>% 
#     sum()) * 100) %>% 
#   round()
# # percent asian in 0-1 and 1-2 tiers
# ((bay_gwthreats_race %>% 
#   filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[1:2]) %>% 
#   filter(race == "Asian Alone") %>% 
#   pull(estimate) %>% 
#   sum()) /
#   (bay_gwthreats_race %>% 
#     filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[1:2]) %>% 
#     pull(estimate) %>% 
#     sum()) * 100) %>% 
#   round()
# # percent black in 0-1 and 1-2 tiers
# ((bay_gwthreats_race %>% 
#   filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[1:2]) %>% 
#   filter(race == "Black or African American") %>% 
#   pull(estimate) %>% 
#   sum()) /
#   (bay_gwthreats_race %>% 
#     filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[1:2]) %>% 
#     pull(estimate) %>% 
#     sum()) * 100) %>% 
#   round()
# # percent black in 100-550 tier
# ((bay_gwthreats_race %>% 
#   filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[3]) %>% 
#   filter(race == "Black or African American") %>% 
#   pull(estimate) %>% 
#   sum()) /
#   (bay_gwthreats_race %>% 
#     filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[3]) %>% 
#     pull(estimate) %>% 
#     sum()) * 100) %>% 
#   round()
# # percent white in 100-550 tier
# ((bay_gwthreats_race %>% 
#   filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[3]) %>% 
#   filter(race == "White Alone") %>% 
#   pull(estimate) %>% 
#   sum()) /
#   (bay_gwthreats_race %>% 
#     filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[3]) %>% 
#     pull(estimate) %>% 
#     sum()) * 100) %>% 
#   round()
# # percent asian in 100-550 tier
# ((bay_gwthreats_race %>% 
#   filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[3]) %>% 
#   filter(race == "Asian Alone") %>% 
#   pull(estimate) %>% 
#   sum()) /
#   (bay_gwthreats_race %>% 
#     filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[3]) %>% 
#     pull(estimate) %>% 
#     sum()) * 100) %>% 
#   round()
# # likelyhood of non-white in 100-550 tier
# ((((bay_gwthreats_race %>% 
#   filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[3]) %>% 
#   filter(race != "White Alone") %>% 
#   pull(estimate) %>% 
#   sum()) /
#   (bay_gwthreats_race %>% 
#     filter(gwthreats_tier %in% bay_gwthreats_race$gwthreats_tier[3]) %>% 
#     pull(estimate) %>% 
#     sum())) / (sum(bay_race_total$estimate[1:6])/sum(bay_race_total$estimate)) - 1) * 100) %>% 
#   round()
```

### Equity Charts Comments

In the Bay Area, the total population is 57% White, 24% Asian, and 6% Black or African American. For the lowest two tiers of groundwater threat score tier of 0 to 1 and 1 to 2, 60% is White, 27% Asian, and 5% Black or African American. Although the differences from the total population race proportions are not very large, White and Asian populations are slightly more numerous and the Black or African American population is slightly smaller. However, the population in the extreme tier with scores between 100 and 550 is 12% Black or African American, which is double the total population percentage. Meanwhile, the extreme tier is 24% Asian, which is equal to the total population percentage, and 49% White, which of course is lower than in total population proportions. Ultimately, this means that non-White individuals seem to be 20% more likely to be in a census tract with a Groundwater Threat Score between 100-550 than expected.

### GAMA Groundwater Contaminant Map

```{r}
leafletOutput("gamaMap")
```


```{r, context = "server"}
observeEvent({
  input$time 
  input$chemical
  }, {
    
  lowBoundDate <-
    switch(
      input$time,
      # "2005-2010" = as.Date("01/01/2005", format = "%m/%d/%Y"),
      # "2010-2015" = as.Date("01/01/2010", format = "%m/%d/%Y"),
      "2015-2017" = as.Date("01/01/2015", format = "%m/%d/%Y"),
      "2017-2019" = as.Date("01/01/2017", format = "%m/%d/%Y"),
      "2019-Present" = as.Date("01/01/2019", format = "%m/%d/%Y")
    )
    
  highBoundDate <-
    switch(
      input$time,
      # "2005-2010" = as.Date("01/01/2011", format = "%m/%d/%Y"),
      # "2010-2015" = as.Date("01/01/2016", format = "%m/%d/%Y"),
      "2015-2017" = as.Date("01/01/2017", format = "%m/%d/%Y"),
      "2017-2019" = as.Date("01/01/2019", format = "%m/%d/%Y"),
      "2019-Present" = as.Date("01/01/2030", format = "%m/%d/%Y")
    )
  
  chem_code <-
    switch(
      input$chemical,
      "Ammonia" = "NH3NH4N",
      "Benzene" = "BZ", 
      "Cadmium" = "CD",
      "Chromium" = "CR",
      "Lead" = "PB",
      "Naphthalene" = "NAPH",
      "Nitrate" = "NO3N",
      "Nitrite" = "NO2",
      "Orthophosphate" = "PORTHO",
      "Toluene" = "BZME",
      "Uranium" = "U",
      "Xylenes" = "XYLENES"
    )
  
  nice_units <-
    switch(
      input$chemical,
      "Ammonia" = "mg/L",
      "Benzene" = "μg/L", 
      "Cadmium" = "μg/L",
      "Chromium" = "μg/L",
      "Lead" = "μg/L",
      "Naphthalene" = "μg/L",
      "Nitrate" = "mg/L",
      "Nitrite" = "mg/L",
      "Orthophosphate" = "mg/L",
      "Toluene" = "μg/L",
      "Uranium" = "pCi/L",
      "Xylenes" = "μg/L"
    )
  
  ca_gama_time <-
    ca_gama %>% 
    mutate(
      DATE = as.Date(DATE, format = "%m/%d/%Y")
    ) %>% 
    filter(
      DATE > lowBoundDate & DATE < highBoundDate
    ) %>% 
    select(
      WELL.ID,
      CHEMICAL,
      RESULTS,
      UNITS,
      DATE,
      LATITUDE,
      LONGITUDE
    ) %>% 
    filter(
      CHEMICAL == chem_code
    ) %>% 
    st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326)  
  
  ca_chemical_pal <- colorNumeric(
  palette = "YlOrRd",
  domain = ca_gama_time$RESULTS
  )
  
  output$gamaMap <- renderLeaflet({
    leaflet() %>% 
      addPolygons(
        data = ca_counties,
        color = "#444444", weight = 1, smoothFactor = 0.5,
        opacity = 0.4, fillOpacity = 0,
        highlightOptions = highlightOptions(color = "white", weight = 2,
          bringToFront = F)
      ) %>% 
      addProviderTiles(providers$CartoDB.Positron) %>%
      addCircleMarkers(
        data = ca_gama_time,
        color = ~ca_chemical_pal(RESULTS),
        label = ~paste0(RESULTS, " ", nice_units,", measured on: ", DATE),
        radius = ~sqrt(RESULTS)
      ) %>% 
      addLegend(
        data = ca_gama_time,
        pal = ca_chemical_pal,
        values = ~RESULTS,
        title = paste0(input$chemical, " Concentrations at USGS Wells in California from ", input$time)
      )
  })
})
```

### GAMA Map Comments

To continue the investigation of groundwater trends, the map above plots the concentrations of key groundwater contaminants at wells in California. The data is sourced from the Groundwater Ambient Monitoring and Assessment Program (GAMA) datasets, which includes well logs from the USGS which track chemical concentrations found in thousands of groundwater sources. Well data comes with the concentrations of 195 different contaminants as well as the date, time and coordinates of the well measurement. The interactive map allows for the time period and one of ten key contaminants to be selected.

### Conclusion
While the Central Valley's large agricultural presence causes groundwater contamination to be at the forefront of scientific attention in the region, the figures above seem to indicate that comparable levels of contaminants and threats are present in certain regions of the Bay Area as well. Further investigations on the census tracts that have been identified with unusually high groundwater contaminant concentrations and threat scores can be instrumental in combating potential health concerns. Moreover, it is evident that racial inequities exist with non-White populations being 20% more likely to face extreme groundwater threat conditions. There are various avenues to pursue further analysis in the realm of groundwater, such as the temporal changes of contaminant levels and the effects of point source and non-point source pollutants on the contaminant levels of nearby aquifers.

