---
title: "vivekvajipey_A1"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = F, warnings = F, messages = F)

# Suppress summarise info
options(dplyr.summarise.inform = FALSE)

library(tidyverse)
library(plotly)
library(gapminder)
```

```{r include=FALSE}
years <- 2017:2021
quarters <- 1:4
type <- "Electric"

pge_2017to2021_elec <- NULL

for(year in years) {
  
  if (year == 2021) {
    quarters <- 1:2
  }
  
  for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  
  temp <- read_csv(filename)
  
  if (year == 2017 & quarter == 4) {
    temp <- temp %>% 
      filter(
        MONTH != 9
      )
  }
  
  pge_2017to2021_elec <- rbind(pge_2017to2021_elec,temp)

  saveRDS(pge_2017to2021_elec, "pge_2017to2021_elec.rds")
  }
}
```

```{r include=FALSE}
years <- 2017:2021
quarters <- 1:4
type <- "Gas"

pge_2017to2021_gas <- NULL

for(year in years) {
  
  if (year == 2021) {
    quarters <- 1:2
  }
  
  for(quarter in quarters) {
  
  filename <- 
    paste0(
      "pge/PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )

  print(filename)
  
  temp <- read_csv(filename)
  
  if (year == 2017 & quarter == 4) {
    temp <- temp %>% 
      filter(
        MONTH != 9
      )
  }
  
  pge_2017to2021_gas <- rbind(pge_2017to2021_gas,temp)

  saveRDS(pge_2017to2021_gas, "pge_2017to2021_gas.rds")
  }
}
```

```{r include=FALSE}
pge_res_elec <-
  pge_2017to2021_elec %>% 
  filter(
    CUSTOMERCLASS %in% 
      c(
        "Elec- Residential"
      )
  ) %>% 
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  ) %>% 
  mutate(TOTALKBTU = TOTALKWH * 3.4121416416) %>% 
  select(
    -c(YEAR, COMBINED, AVERAGEKWH)
  ) %>% 
  group_by(DATE, CUSTOMERCLASS) %>%
  summarize(
    TOTALKBTU =
      sum(
        TOTALKBTU,
        na.rm = T
      ),
    TOTALCUSTOMERS =
      sum(
        TOTALCUSTOMERS,
        na.rm = T
      )
  ) %>%
  mutate(
    AVERAGEKBTU =
      TOTALKBTU/TOTALCUSTOMERS
  )
```

```{r include=FALSE}
pge_res_gas <-
  pge_2017to2021_gas %>% 
  filter(
    CUSTOMERCLASS %in% 
      c(
        "Gas- Residential"
      )
  ) %>% 
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  ) %>% 
  mutate(TOTALKBTU = TOTALTHM * 99.9761) %>% 
  select(
    -c(YEAR, COMBINED, AVERAGETHM)
  ) %>% 
  group_by(DATE, CUSTOMERCLASS) %>%
  summarize(
    TOTALKBTU =
      sum(
        TOTALKBTU,
        na.rm = T
      ),
    TOTALCUSTOMERS =
      sum(
        TOTALCUSTOMERS,
        na.rm = T
      )
  ) %>%
  mutate(
    AVERAGEKBTU =
      TOTALKBTU/TOTALCUSTOMERS
  )

pge_res_gas
```

```{r include=FALSE}
pge_residential_final = rbind(pge_res_elec, pge_res_gas)
pge_residential_final
```

```{r}
ggplot(pge_residential_final, 
       aes(
         x = DATE,
         y = TOTALKBTU / 1e9
       )) +
  geom_line(
    aes(
      color = CUSTOMERCLASS,
      linetype = CUSTOMERCLASS
    )
  ) +
   scale_color_manual(values = c("darkred", "steelblue")) +
  labs(
      x = "",
      y = "Total kBTUs (billions)",
      title = "PG&E Territory Monthly Residential Electricity and Gas Usage, 2017 to 2021",
    )
```

```{r include=FALSE}
pge_com_elec <-
  pge_2017to2021_elec %>% 
  filter(
    CUSTOMERCLASS %in% 
      c(
        "Elec- Commercial"
      )
  ) %>% 
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  ) %>% 
  mutate(TOTALKBTU = TOTALKWH * 3.4121416416) %>% 
  select(
    -c(YEAR, COMBINED, AVERAGEKWH)
  ) %>% 
  group_by(DATE, CUSTOMERCLASS) %>%
  summarize(
    TOTALKBTU =
      sum(
        TOTALKBTU,
        na.rm = T
      ),
    TOTALCUSTOMERS =
      sum(
        TOTALCUSTOMERS,
        na.rm = T
      )
  ) %>%
  mutate(
    AVERAGEKBTU =
      TOTALKBTU/TOTALCUSTOMERS
  )
```

```{r include=FALSE}
pge_com_gas <-
  pge_2017to2021_gas %>% 
  filter(
    CUSTOMERCLASS %in% 
      c(
        "Gas- Commercial"
      )
  ) %>% 
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  ) %>% 
  mutate(TOTALKBTU = TOTALTHM * 99.9761) %>% 
  select(
    -c(YEAR, COMBINED, AVERAGETHM)
  ) %>% 
  group_by(DATE, CUSTOMERCLASS) %>%
  summarize(
    TOTALKBTU =
      sum(
        TOTALKBTU,
        na.rm = T
      ),
    TOTALCUSTOMERS =
      sum(
        TOTALCUSTOMERS,
        na.rm = T
      )
  ) %>%
  mutate(
    AVERAGEKBTU =
      TOTALKBTU/TOTALCUSTOMERS
  )
```

```{r include=FALSE}
pge_commercial_final = rbind(pge_com_elec, pge_com_gas)
pge_commercial_final
```

```{r}
ggplot(pge_commercial_final, 
       aes(
         x = DATE,
         y = TOTALKBTU / 1e9
       )) +
  geom_line(
    aes(
      color = CUSTOMERCLASS,
      linetype = CUSTOMERCLASS
    )
  ) +
   scale_color_manual(values = c("darkred", "steelblue")) +
  labs(
      x = "",
      y = "Total kBTUs (billions)",
      title = "PG&E Territory Monthly Commercial Electricity and Gas Usage, 2017 to 2021",
    )
```

```{r include=FALSE}
pge_commercial_covid <-
  pge_commercial_final %>% 
    .[25:54,] %>% 
  rbind(
    pge_commercial_final %>% 
    .[79:108,]
  )

yearly_sums_com <-
  c(
    sum(pge_commercial_covid$AVERAGEKBTU[1:12]), 
    sum(pge_commercial_covid$AVERAGEKBTU[31:42]), 
    sum(pge_commercial_covid$AVERAGEKBTU[13:24]), 
    sum(pge_commercial_covid$AVERAGEKBTU[43:54])
  )

pge_covid_com_sums <- data.frame( 
  Year = c("2019", "2019", "2020", "2020"),
  AveragekBTU = yearly_sums_com,
  CustomerClass = c("Elec- Commercial", "Gas- Commercial", "Elec- Commercial", "Gas- Commercial")
)
```

```{r}
pge_covid_com_sums_chart <-
  pge_covid_com_sums %>% 
  ggplot() +
  geom_bar(
    aes(
      x = Year,
      y = AveragekBTU / 1e3,
      fill = CustomerClass
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Year",
    y = "kBTU per customer (thousands)",
    title = "PG&E Territory Average Yearly Commercial Electricity and Gas Usage, 2019 and 2020",
    fill = "Energy Type"
  )

pge_covid_com_sums_chart %>% ggplotly()
```

```{r}
pge_residential_covid <-
  pge_residential_final %>% 
    .[25:54,] %>% 
  rbind(
    pge_residential_final %>% 
    .[79:108,]
  )
  
yearly_sums_res <-
  c(
    sum(pge_residential_covid$AVERAGEKBTU[1:12]), 
    sum(pge_residential_covid$AVERAGEKBTU[31:42]), 
    sum(pge_residential_covid$AVERAGEKBTU[13:24]), 
    sum(pge_residential_covid$AVERAGEKBTU[43:54])
  )

pge_covid_res_sums <- data.frame(
  Year = c("2019", "2019", "2020", "2020"),
  AveragekBTU = yearly_sums_res,
  CustomerClass = c("Elec- Residential", "Gas- Residential", "Elec- Residential", "Gas- Residential")
)
```

```{r}
pge_covid_res_sums_chart <-
  pge_covid_res_sums %>% 
  ggplot() +
  geom_bar(
    aes(
      x = Year,
      y = AveragekBTU / 1e3,
      fill = CustomerClass
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Year",
    y = "kBTU per ",
    title = "PG&E Territory Yearly Residential Electricity and Gas Usage, 2019 and 2020",
    fill = "Energy Type"
  )

pge_covid_res_sums_chart %>% ggplotly()
```
From 2019 to 2020, the average commercial energy consumption decreased while the residential energy consumption increased in the PG&E Territory.  This change may be attributable to a large portion of the population spending more time in their residences during the COVID-19 pandemic in 2020 than in 2019. In the bar charts, average energy consumption is plotted rather than total energy consumption to better isolate the effect of the pandemic from changes in the number of PG&E customers.

The data used in these charts are from PG&E's public datasets, which report customer usage data for gas in therms and electricity in kWh, which have been converted to kBTUs for ease of comparison. The data is reported by ZIP code, so the charts above include the regions that receive energy from PG&E in the Bay Area. The datasets satisfy the following rules for public aggregation of data: "a minimum of 100 Residential customers; a minimum of 15 Non-Residential customers, with no single Non-Residential customer accounting for more than 15% of the total consumption. If aggregation is not met, the consumption will be combined with a neighboring ZIP code until the aggregation requirements are met." The 2017 Quarter 4 Electricity Usage dataset happened to include September values however, these values were redundant with the 2017 Quarter 3 report so the September values of Q4 were filtered out.