---
title: "vivekvajipey_A5"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warnings = F, messages = F)

library(tidyverse)
library(sf)
library(tigris)
library(leaflet)
library(censusapi)
library(mapview)
library(readxl)

Sys.setenv(CENSUS_KEY="67343486046933b2d74f13923ac153228348f817")
```

```{r include=FALSE}
temp <- tempfile()
download.file("https://oehha.ca.gov/media/downloads/calenviroscreen/document/calenviroscreen40resultsdatadictionaryf2021.zip",destfile = temp)

ces4 <- read_excel("calenviroscreen40resultsdatadictionary_F_2021.xlsx", sheet = "CES4.0FINAL_results")
```

```{r include=FALSE}
bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

ca_tracts <- tracts("CA", cb = T, progress_bar = F)

ces4_bay_pm25 <-
  ces4 %>% 
  filter(`California County` %in% bay_county_names) %>% 
  select(`Census Tract`, PM2.5) %>% 
  left_join(
    ca_tracts %>% 
      transmute(GEOID = as.numeric(GEOID)), 
    by = c("Census Tract" = "GEOID")
  ) %>% 
  st_as_sf()
```

```{r}
pm25_pal <- colorNumeric(
  palette = "Reds",
  domain = ces4_bay_pm25$PM2.5
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = ces4_bay_pm25,
    fillColor = ~pm25_pal(PM2.5),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~PM2.5
  ) %>% 
  addLegend(
    data = ces4_bay_pm25,
    pal = pm25_pal,
    values = ~PM2.5,
    title = "Annual mean concentration of PM2.5 in µg/m^3 from 2015 to 2017"
  )
```

```{r include=FALSE}
ces4_bay_asthma <-
  ces4 %>% 
  filter(`California County` %in% bay_county_names) %>% 
  select(`Census Tract`, Asthma) %>% 
  left_join(
    ca_tracts %>% 
      transmute(GEOID = as.numeric(GEOID)), 
    by = c("Census Tract" = "GEOID")
  ) %>% 
  st_as_sf()
```

```{r include=FALSE}
asthma_pal <- colorNumeric(
  palette = "Reds",
  domain = ces4_bay_asthma$Asthma
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = ces4_bay_asthma,
    fillColor = ~asthma_pal(Asthma),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~Asthma
  ) %>% 
  addLegend(
    data = ces4_bay_asthma,
    pal = asthma_pal,
    values = ~Asthma,
    title = "Age-adjusted rate of emergency department visits for asthma per 10,000"
  )
```

The map of PM2.5 in the Bay Area shows the annual mean concentration of PM2.5 in µg/m3 from 2015 to 2017 at the census tract level. The map of asthma in the Bay Area shows the age-adjusted rate of emergency department visits for asthma per 10,000 (averaged over 2015-2017). Two areas that stick out as facing from both high concentrations of PM2.5 and high rates of asthma-caused ED visits are Vallejo and Anitoch.

```{r include=FALSE}
ces4_bay_pm25_asthma <-
  ces4 %>% 
  filter(`California County` %in% bay_county_names) %>% 
  select(`Census Tract`, PM2.5, Asthma) %>% 
  left_join(
    ca_tracts %>% 
      transmute(GEOID = as.numeric(GEOID)), 
    by = c("Census Tract" = "GEOID")
  ) %>% 
  st_as_sf()
```


```{r}
ggplot(
  data = ces4_bay_pm25_asthma,
  aes(
      x = PM2.5,
      y = Asthma
    )
) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Asthma vs PM2.5 in the Bay Area")

```

The line of best fit does not seem to fit the data very well, due to the rather jumbled and unevenly distributed nature of the data points. Between 8 and 9 µg/m^3, there is a large cluster of points that greatly exceed the asthma prevalence of the best fit line.


```{r}
pm25_asthma_model <- lm(Asthma ~ PM2.5, ces4_bay_pm25_asthma)

summary(pm25_asthma_model)
```
Based on this linear regression model, an increase of 1 µg/m3 of annual mean concentration of PM2.5 is correlated with an increase of 19.862 emergency department visits for asthma per 10,000. The standard error of this slope value is 1.534.

The r-squared value is approximately 0.096, indicating that the variation in annual mean PM2.5 concentrations explains 9.6% of the variation in asthma-caused emergency department vists.

```{r}
plot(density(residuals(pm25_asthma_model)))
```
The mean of the residual is quite a bit less than zero and the entire density curve seems to be skewed to the left. These conditions indicate that the regression for the data cannot be interpretted meaningfully.

```{r}
ggplot(
  data = ces4_bay_pm25_asthma,
  aes(
      x = PM2.5,
      y = log(Asthma)
    )
) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("log(Asthma) vs PM2.5 in the Bay Area")
```
This scatter plot also exhibits the vertical cluster of data points between 8 and 9 µg/m^3, however compared to the linear model, this log model seems to fit the data points more closely.

```{r}
pm25_asthma_model_log <- lm(log(Asthma) ~ PM2.5, ces4_bay_pm25_asthma)

summary(pm25_asthma_model_log)
```
Based on this linear regression model, an increase of 1 µg/m3 of annual mean concentration of PM2.5 is correlated with an increase of 0.356% in the number of emergency department visits for asthma per 10,000. The standard error of this slope value is 0.6566.

The r-squared value is approximately 0.10, indicating that the variation in annual mean PM2.5 concentrations explains 10% of the variation in asthma-caused emergency department vists.

```{r}
plot(density(residuals(pm25_asthma_model_log)))
```
This distribution of residuals is significantly more normal than that of the linear model, with the mean being very close to zero and the overall curve being neither skewed right nor left. The clear distinction between the curve and a standard normal distribution is the dip at the mean with two peaks to the left and right.
