---
title: "vivekvajipey_A4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warnings = F, messages = F)

library(tidyverse)
library(sf)
library(tigris)
library(leaflet)
library(censusapi)
library(mapview)
library(readxl)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5",
    type = "variables"
  )
```

```{r include=FALSE}
census_race_categories <- 
  c(
    "White Alone",
    "Black or African American",
    "American Indian and Alaska Native Alone",
    "Asian Alone",
    "Native Hawaiian and Other Pacific Islander Alone",
    "Some Other Race Alone",
    "Two or More Races"
  )

sc_edu_race <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "county:085",
      regionin = "state:06",
      vars = paste0("group(C15002", LETTERS[x],")")
    ) %>%
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,NA,"education"),
        sep = "!!"
      ) %>% 
      filter(!is.na(education)) %>% 
      mutate(race = census_race_categories[x])
  })
```

```{r, echo = F, warnings = F, messages = F}
sc_edu_race %>% 
  group_by(education, race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = education %>% factor(levels = unique(sc_edu_race$education)),
      y = estimate,
      fill = race %>% factor(levels = unique(sc_edu_race$race))
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Educational attainment",
    y = "Proportion of population 25 years and older",
    title = "Santa Clara County education attainment by race",
    fill = "Race"
  ) +
  theme(axis.text.x = element_text(angle = 90)) +
  coord_flip()
```

```{r, echo = F, warnings = F, messages = F}
sc_race_total <-
  sc_edu_race %>% 
  group_by(race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(education = "Total")

sc_edu_race %>% 
  group_by(education, race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  rbind(sc_race_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = education %>% factor(levels = rev(c("Total",unique(sc_edu_race$education)))),
      y = estimate,
      fill = race %>% factor(levels = rev(unique(sc_edu_race$race)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Educational attainment",
    y = "Proportion of population 25 years and older",
    title = "Santa Clara County education attainment by race",
    fill = "Race"
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )  +
  guides(
    fill = guide_legend(
      reverse = T
    )
  )
```

```{r include=FALSE}
# Percent of population non-white
((sum(sc_race_total$estimate[1:6])/sum(sc_race_total$estimate))*100) %>% round()
```

```{r include=FALSE}
# Percent of population asian alone
((sum(sc_race_total$estimate[2])/sum(sc_race_total$estimate))*100) %>% round()
```

```{r include=FALSE}
# Percent of population black or african american alone
((sum(sc_race_total$estimate[3])/sum(sc_race_total$estimate))*100) %>% round()
```

```{r include=FALSE}
# Percent non-white of population with high school graduate education or less
((sc_edu_race %>% 
  filter(education %in% sc_edu_race$education[1:2]) %>% 
  filter(race != "White Alone") %>% 
  pull(estimate) %>% 
  sum()) /
  (sc_edu_race %>% 
    filter(education %in% sc_edu_race$education[1:2]) %>% 
    pull(estimate) %>% 
    sum()) * 100) %>% 
  round()
```

```{r include=FALSE}
# Percent non-white of population with less than a high school diploma
((sc_edu_race %>% 
  filter(education %in% sc_edu_race$education[1]) %>% 
  filter(race != "White Alone") %>% 
  pull(estimate) %>% 
  sum()) /
  (sc_edu_race %>% 
    filter(education %in% sc_edu_race$education[1]) %>% 
    pull(estimate) %>% 
    sum()) * 100) %>% 
  round()
```

```{r include=FALSE}
# percent by which non-whites are more likely to have high school education or less
((((sc_edu_race %>% 
  filter(education %in% sc_edu_race$education[1]) %>% 
  filter(race != "White Alone") %>% 
  pull(estimate) %>% 
  sum()) /
  (sc_edu_race %>% 
    filter(education %in% sc_edu_race$education[1]) %>% 
    pull(estimate) %>% 
    sum())) / (sum(sc_race_total$estimate[1:6])/sum(sc_race_total$estimate)) - 1) * 100) %>% 
  round()
```

```{r include=FALSE}
# Percent black of population with bachelor's or higher
((sc_edu_race %>% 
  filter(education %in% sc_edu_race$education[4]) %>% 
  filter(race == "Black or African American") %>% 
  pull(estimate) %>% 
  sum()) /
  (sc_edu_race %>% 
    filter(education %in% sc_edu_race$education[4]) %>% 
    pull(estimate) %>% 
    sum()) * 100) %>% 
  round()
```

In Santa Clara County, 54% of the population over 25 is non-white and 57% of the population over 25 with a high school graduate education or lower is non-white, which fairly close. However, the population over 25 with an education strictly less than a high school diploma is 64% non-white, and non-white individuals seem to be 19% more likely to not have a high school education than expected. For many jobs, a baseline of a high school education is required, which narrows the set of available opportunities for the non-white population in this lowest education tier. The data represent the education of individuals older than 25 and many individuals who complete high school and a potential associate's/bachelor's degree before turning 25, so these data generally represent the highest education to be attained by the individual. It may be insightful to consider the effects of wealth and income on the educational disparities between races, especially since much of Santa Clara county is wealthy.

```{r include=FALSE}
census_eth_categories <- 
  c(
    "White Alone, Not Latinx",
    "Latinx"
  )

sc_edu_eth <-
  1:2 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "county:085",
      regionin = "state:06",
      vars = paste0("group(C15002",LETTERS[x+7],")")
    ) %>%
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,NA,"education"),
        sep = "!!"
      ) %>% 
      filter(!is.na(education)) %>% 
      mutate(ethnicity = census_eth_categories[x])
  }) %>% 
  group_by(ethnicity, education) %>% 
  summarize(estimate = sum(estimate))
```

```{r}
census_edu_categories <- 
  c(
    "Total",
    "No schooling completed",
    "Nursery completed",
    "Kindergarten",
    "1st grade",
    "2nd grade",
    "3rd grade",
    "4th grade",
    "5th grade",
    "6th grade",
    "7th grade",
    "8th grade",
    "9th grade",
    "10th grade",
    "11th grade",
    "12th grade, no diploma",
    "Regular high school diploma",
    "GED or alternative credential",
    "Some college, less than 1 year",
    "Some college, 1 or more years, no degree",
    "Associate's degree",
    "Bachelor's degree",
    "Master's degree",
    "Professional school degree",
    "Doctorate degree"
  )

sc_edu_total <-
  getCensus(
    name = "acs/acs5",
    vintage = 2019,
    region = "county:085",
    regionin = "state:06",
    vars = "group(B15003)"
  ) %>%
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "education",
    values_to = "estimate"
  ) %>% 
  mutate(
    education = census_edu_categories
  ) %>% 
  subset(select = -c(county))
```

```{r}
education_levels <- c("Less than high school diploma", "High school graduate (includes equivalency)", "Some college or associate's degree", "Bachelor's degree or higher", "Total")

sc_edu_total <-
  sc_edu_total %>%
  add_row(education = education_levels[1], estimate = sum(sc_edu_total$estimate[2:16])) %>%  
  add_row(education = education_levels[2], estimate = sum(sc_edu_total$estimate[17:18])) %>% 
  add_row(education = education_levels[3], estimate = sum(sc_edu_total$estimate[19:21])) %>% 
  add_row(education = education_levels[4], estimate = sum(sc_edu_total$estimate[22:25])) %>% 
  filter(
    education %in% c("Less than high school diploma", "High school graduate (includes equivalency)", "Some college or associate's degree", "Bachelor's degree or higher")
  ) %>% 
  group_by(education) %>% 
  summarize(estimate = sum(estimate))
```

```{r echo = F, warnings = F, messages = F}
sc_edu_white_and_latinx <-
  sc_edu_eth %>% 
  group_by(
    education
  ) %>% 
  summarize(
    estimate = sum(estimate)
  )

sc_edu_other_eth <- 
  sc_edu_total %>% 
  mutate(
    estimate = sc_edu_total$estimate - sc_edu_white_and_latinx$estimate,
    ethnicity = "Not White, Not Latinx"
  )

sc_edu_other_eth <- sc_edu_other_eth[,c(3, 1,2)]
  
sc_edu_eth <- 
  rbind(
    sc_edu_eth,
    sc_edu_other_eth
  )
```

```{r, }
sc_edu_eth %>% 
  ggplot() +
  geom_bar(
    aes(
      x = education %>% factor(levels = education_levels),
      y = estimate,
      fill = ethnicity %>% factor(levels = unique(sc_edu_eth$ethnicity))
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Educational attainment",
    y = "Proportion of population 25 years and older",
    title = "Santa Clara County education attainment by ethnicity",
    fill = "Ethnicity"
  ) +
  theme(axis.text.x = element_text(angle = 90)) +
  coord_flip()
```

```{r echo = F, warnings = F, messages = F}
sc_eth_total <-
  sc_edu_eth %>% 
  group_by(ethnicity) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(education = "Total")

sc_edu_eth %>% 
  group_by(education, ethnicity) %>% 
  summarize(estimate = sum(estimate)) %>% 
  rbind(sc_eth_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = education %>% factor(levels = education_levels),
      y = estimate,
      fill = ethnicity %>% factor(levels = rev(unique(sc_edu_eth$ethnicity)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Educational attainment",
    y = "Proportion of population 25 years and older",
    title = "Santa Clara County education attainment by ethnicity",
    fill = "Ethnicity"
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )  +
  guides(
    fill = guide_legend(
      reverse = T
    )
  )
```
...
```{r include=FALSE}
# Percent of population asian alone
((sum(sc_eth_total$estimate[3])/sum(sc_eth_total$estimate))*100) %>% round()
```

```{r include=FALSE}
# Percent of population for each ethnicity
((sum(sc_race_total$estimate[1])/sum(sc_race_total$estimate))*100) %>% round()
((sum(sc_race_total$estimate[2])/sum(sc_race_total$estimate))*100) %>% round()
((sum(sc_race_total$estimate[3])/sum(sc_race_total$estimate))*100) %>% round()
```

```{r include=FALSE}
# Percent latinx of population with strictly less than high school diploma
((sc_edu_eth %>% 
  filter(education %in% sc_edu_eth$education[3]) %>% 
  filter(ethnicity == "Latinx") %>% 
  pull(estimate) %>% 
  sum()) /
  (sc_edu_race %>% 
    filter(education %in% sc_edu_eth$education[3]) %>% 
    pull(estimate) %>% 
    sum()) * 100) %>% 
  round()
```
```{r include=FALSE}
# Percent white of population with strictly less than high school diploma
((sc_edu_eth %>% 
  filter(education %in% sc_edu_eth$education[7]) %>% 
  filter(ethnicity == "White Alone, Not Latinx") %>% 
  pull(estimate) %>% 
  sum()) /
  (sc_edu_race %>% 
    filter(education %in% sc_edu_eth$education[3]) %>% 
    pull(estimate) %>% 
    sum()) * 100) %>% 
  round()
```

```{r include=FALSE}
# Percent latinx of population with high school graduate education or less
((sc_edu_eth %>% 
  filter(education %in% sc_edu_eth$education[2:3]) %>% 
  filter(ethnicity == "Latinx") %>% 
  pull(estimate) %>% 
  sum()) /
  (sc_edu_race %>% 
    filter(education %in% sc_edu_eth$education[2:3]) %>% 
    pull(estimate) %>% 
    sum()) * 100) %>% 
  round()
```

In Santa Clara County, 21% of the population over 25 is Latinx, 43% of the population over 25 is neither white nor Latinx and 35% of the population over 25 is white alone and not Latinx. However, the population over 25 with an education of a high school degree or lower is 47% Latinx. The statistics are even further from the overall population percentages for the population with an education strictly less than a high school diploma, which is 56% Latinx and only 10% white alone and not Latinx.
